Formal Verification of Real-Time Data Processing of the LHC Beam Loss Monitoring System: A Case Study

### 2011-03-24 fmics/paper.tex
remove trailing whitespace
merge changes from dropbox

The SRS component has a complex structure and the correctness of its behaviour is critical for safe and productive use of the LHC.
Any error in the SRS implementation would either compromise the availability of the LHC (unnecessary request for beam dump) or safety (not triggering a necessary beam dump).
The current approach for analysing the SRS implementation is simulation of its behavior on sample streams of input~\cite{}.
(one more sentence needs to be added on the current verification process)

In this paper, we describe a formal verification approach based on logical deduction for analysing the SRS implementation.
(need to add few sentences explaining in a very high level the verification approach/strategy)

Compared to test-based methods, like simulation, formal methods not only offer a much higher confidence in correctness of a system's behavior, but also help improve our understanding of its specification.
One of the challenges in pursuing a formal verification approach for SRS was capturing the intricate details of the system's specification via experiment and refinement with a team with different backgrounds and expertise.
Our confidence in the SRS design as a result of this effort ultimately rests upon our deep understanding of why the design is correct rather than the fact that we obtained ``Theorem Proved'' as the final output of a software tool at the end of a long and intense effort.
In particular, our use of mechanized logical deduction was a highly iterative process that incrementally refined our understanding of (1) the implementation (2) the intended behavior and (3) the ``white board level'' argument or explanation for why the implementation achieves its intended behaviors.
The most important use of HOL4 was its role as an ``implacable skeptic'' that compelled us to really understand the details~\cite{rushby}.

@@

In order to construct long moving windows, it is required to keep long histories of received data. The technique employed to reach long integration periods with relatively small in length shift registers uses consecutive storage of sums of the received values. Instead of storing all values needed for the sum, this technique stores successively parts of the total sum using only a fraction of the otherwise needed memory space. This technique works by feeding the sum of one shift register's contents, every time its contents become completely updated, to the input of another shift register (see Figure~\ref{fig:SRS-basic}). By cascading more of these elements, very long moving windows could be constructed with significantly small memory space. The above scheme is the basis for the SRS implementation in BLETC.

@@

Our goal is to build a generic model of the SRS structure, and to prove that it calculates approximate running sums of counts within suitable error margins.
Any statement in formal logic is necessarily precise, and one of our challenges has been characterizing our goal precisely.
A sketch of the desired statement is \[\forall{n}.\,\text{\textsf{RS} }n=\text{sum for output $n$}\pm\text{acceptable error}\]
Although Figure~\ref{fig:srs} would suggest we only care about $0\leq{n}\leq11$, if we can define and prove something for all values of $n$, we obtain a generic model.

To prove the equation above, we need to formalize all of its terms.
First, we make two items explicit: that the equation is to be understood as holding at each time step, and that the terms depend on the stream of input counts over time.
With these refinements, our statement becomes \[\forall{D\,n\,t}.\,\mathsf{RS}\;D\;n\;t=\text{sum, given input stream $D$, for output $n$ at time $t$}\pm\text{acceptable error}\]
(The input stream is a function of time: $D\;t$ is the input value to the whole SRS structure at time $t$.)

To formalize the term on the left hand side, we need to build a logical model of the SRS structure.
Our work proceeded in two phases, starting with a model of a simplified structure and proceeding to a more realistic model.
In both cases, we represent each piece of the SRS structure that has a value---for example, each cell in each shift register---as a function in HOL that takes the global input stream and current time step as input and returns the value of the piece of the structure at that time.
Our simplifications in the first phase include
\begin{enumerate}
\item Ignore the middle taps in each slice
\item A more regular arrangement of slices than the actual structure
\item Ignore subtractor and accumulator, by defining the output of a slice as the sum of its cells directly
\item Simplify timing <what? how?>
\item Use natural numbers, rather than finite words, for cell values
\end{enumerate}
In the second phase, we drop the first three simplifications by adding middle taps, a more flexible arrangement of slices, and the subtractor and accumulator to our model.
In both cases, however, the model is simultaneously a simplification and a generalization of the actual structure in the following sense: we model an infinite number of slices, each with an infinite number of shift register cells and taps, simply by letting indices range over the natural numbers without explicitly giving limits.

We describe the more realistic (second phase) model here.
Table~\ref{tab:descriptions} lists the names of the HOL functions we define and their intended meanings, and Figure~\ref{fig:definitions} gives their definitions. 
For our correctness equation above, we only need the last entry in the table, $\mathsf{RS}$; the first two entries, $\mathsf{tap}$ and $\mathsf{input}$, enable us to specify the exact configuration of the structure, and the remaining entries are used as ``glue'' leading up to the definition of $\mathsf{RS}$.
By using these intermediary functions, every definition in Figure~\ref{fig:definitions} is \emph{local} (only talks about a small part of the overall SRS structure) and therefore is easy to verify against its intended meaning.

>>>>

The SRS component has a complex structure and the correctness of its behaviour is critical for safe and productive use of the LHC.
Any error in the SRS implementation would either compromise the availability of the LHC (unnecessary request for beam dump) or safety (not triggering a necessary beam dump).
The current approach for analyzing the SRS implementation is simulation of its behavior on sample streams of input~\cite{}.
(one more sentence needs to be added on the current verification process)

In this paper, we describe a formal verification approach based on logical deduction for analyzing the SRS implementation.
(need to add few sentences explaining in a very high level the verification approach/strategy)

Compared to test-based methods, like simulation, formal methods not only offer a much higher confidence in correctness of a system's behavior, but also help improve our understanding of its specification.
One of the challenges in pursuing a formal verification approach for SRS was capturing the intricate details of the system's specification via experiment and refinement with a team with different backgrounds and expertise.
Our confidence in the SRS design as a result of this effort ultimately rests upon our deep understanding of why the design is correct rather than the fact that we obtained ``Theorem Proved" as the final output of a software tool at the end of a long and intense effort.
In particular, our use of mechanized logical deduction was a highly iterative process that incrementally refined our understanding of (1) the implementation (2) the intended behavior and (3) the ``white board level" argument or explanation for why the implementation achieves the intended behaviors.
The most important use of HOL4  was its role as an ``implacable skeptic" that compelled us to really understand the details~\cite{rushby}.

@@

In order to construct long moving windows, it is required to keep long histories of received data, referred to as \emph{count} values. The technique employed to reach long integration periods with relatively small in length shift registers uses consecutive storage of sums of the counts. Instead of storing all values needed for the sum, this technique stores successively parts of the total sum using only a fraction of the otherwise needed memory space. This technique works by feeding the sum of one shift register's contents, every time its contents become completely updated, to the input of another shift register (see Figure~\ref{fig:SRS-basic}). By cascading more of these elements, very long moving windows could be constructed with significantly small memory space. The above scheme is the basis for the SRS implementation in BLETC.

@@

Our goal is to build a generic model of the SRS structure, and to prove that it satisfies its specification, i.e. it calculates approximate running sums of received count values within acceptable error margins.
%Any statement in formal logic is necessarily precise, and one of our challenges has been characterizing our goal precisely.
%One of our challenges has been characterizing this goal precisely
Let $\mathsf{output} \, n$ denote the sum of the last $m$ received count values and $\mathsf{RS} \, n$ denote one of the outputs computed by the slices in the structure of SRS which is supposed to compute the sum of the last $m$ received count values with some delay.
A sketch of the desired correctness statement is \[\forall{n} \cdot \, \mathsf{RS} \,n= \mathsf{output} \, n \pm\text{acceptable error}\]
Although Figure~\ref{fig:srs} suggests that in the SRS structure there are only twelve outputs  (i.e. $1\leq{n}\leq11$), we would obtain a more generic result (that is useful in future upgrades of the system) if we prove the correctness of the above statement for all values of $n$.

%To prove the equation above, we need to formalize all of its terms.
In order to make the above correctness statement more precise, we need to include the notion of time.
The count values arrive at the input of the SRS block every 40 micro-seconds that is abstracted by one time step in our logical model.
We formalize the input stream to SRS as a function of time: $D\;t$ denotes the input value to the SRS structure at time $t$.
In addition, the terms in the above statement depend on the stream of input counts over time.
With these refinements, the correctness statement is defines as: \[\forall{D\,n\,t} \cdot \,\mathsf{RS} \,D\,n\,t = \mathsf{output} \,D\,\,n\, t \pm\text{acceptable error}\]

%First, we make two items explicit: that the equation is to be understood as holding at each time step, and that the terms depend on the stream of input counts over time.
%With these refinements, our statement becomes \[\forall{D\,n\,t}.\,\mathsf{RS}\;D\;n\;t=\text{sum, given input stream $D$, for output $n$ at time $t$}\pm\text{acceptable error}\]
%(The input stream is a function of time: $D\;t$ is the input value to the whole SRS structure at time $t$.)

%To formalize the term on the left hand side, we need to build a logical model of the SRS structure.
The first natural step to prove the correctness of the above statement is to build a logical model of the SRS structure.
Our work proceeded in two phases, starting with a model of a simplified structure and proceeding to a model which is closer to the SRS implementation in BLMS.
Our main approach in building the logical model of SRS (in both phases) was to model each building block of the SRS structure that holds a count value--for example, each cell in each shift register--as a function in HOL that takes the input stream and current time step as inputs and returns the value which the corresponding building block holds at that time.

%In both cases, we represent each piece of the SRS structure that has a value---for example, each cell in each shift register---as a function in HOL that takes the global input stream and current time step as input and returns the value of the piece of the structure at that time.

Our simplifications in the first phase include:
\begin{enumerate}
\item Masking the middle taps in each slice,
\item Considering a more regular arrangement of slices than the SRS implementation in BLMS,
\item Masking the subtractors and the accumulators in each slice, instead, defining the output of a slice as the sum of the values in its cells directly,
\item Simplifying timing by ignoring some of the delays that each building block may introduce,
\item Using natural numbers, rather than finite words, for count values
\end{enumerate}

Building a logical model for simplified structure of the SRS helped us improve our understanding of the behavior of the SRS.
{\bf what else did we learn through this process? If there was no particular advantage, I don't think we should mention the simplified version}.

In the second phase, we eliminated the first four simplifications (listed above) by adding middle taps, a more flexible arrangement of slices, the subtractor and accumulator, and considering the delays that each block may introduce.
In both cases, however, the model is simultaneously a simplification and a generalization of the actual structure of SRS in BLMS in the following aspects: we model an infinite number of slices (rather than six slices), each with an infinite number of shift register cells and taps (rather than fixed width shift registers and only two taps), simply by letting indices range over the natural numbers without explicitly giving limits. Such a parameterization would make the model and the analysis applicable to the future upgraded/modified versions of the system.

%%%%% upto here %%%
We describe the more realistic (second phase) model here.
Table~\ref{tab:descriptions} lists the names of the HOL functions we define and their intended meanings, and Figure~\ref{fig:definitions} gives their definitions.
For our correctness equation above, we only need the last entry in the table, $\mathsf{RS}$; the first two entries, $\mathsf{tap}$ and $\mathsf{input}$, enable us to specify the exact configuration of the structure, and the remaining entries are used as ``glue'' leading up to the definition of $\mathsf{RS}$.
By using these intermediary functions, every definition in Figure~\ref{fig:definitions} is \emph{local} (only talks about a small part of the overall SRS structure) and therefore is easy to verify against its intended meaning.
